---
title: "Wishart vs inverse Wishart"
author: "Gerko Vink"
date: "6/13/2020"
output: html_document
---

```{r message = FALSE}
library(MCMCpack)
library(magrittr)
set.seed(123)
```

The question is whether an inverse Wishart distribution's inverse $\mathbf{X}^{-1}$ follows: 

- $\mathcal{W}(\mathbf \Psi^{-1}, \nu)$, or 
- $\mathcal{W}(\mathbf \Psi, \nu)$ (cf. Schaffer, 1997):

To determine which is correct, we start with drawing a $\Sigma$

```{r}
#Make Sigma
Sigma <- c(1, rep(c(.1, .2, .3), 10 )) %>% 
  toeplitz 
```
to draw from the inverse Wishart distribution, with $\nu = 31$
```{r}
# Draw inverse wishart
inv.wish <- riwish(31, Sigma)
```
Then, we create an estimated wishart from the generated inverse Wishart, simply by taking the inverse of the generated inverse Wishart:
```{r}
# Create `estimated` wishart from inverse wishart
est.wish <- solve(inv.wish)
```

To test whether $\mathbf{X}$ follows an inverse Wishart distribution if its inverse $\mathbf{X}^{-1} \sim \mathcal{W}(\mathbf \Psi^{-1}, \nu)$ or if its inverse $\mathbf{X}^{-1} \sim \mathcal{W}(\mathbf \Psi, \nu)$ (cf. Schaffer, 1997):
```{r}
# Draw 2 versions of wishart - one with and one without inverse Sigma
drawn.wish <- rwish(31, solve(Sigma))
drawn.wish2 <- rwish(31, Sigma)
```

We can verify which distribution it follows by making a `qqplot`:
```{r}
# Test whether estimated wishart from inverse wishart conforms to a drawn wishart 
qqplot(est.wish, drawn.wish); abline(coef = c(0,1)) #correct
qqplot(est.wish, drawn.wish2);  abline(coef = c(0,1))
```

### Conclusion
We say $\mathbf{X}$ follows an inverse Wishart distribution, 
denoted as $\mathbf{X}\sim \mathcal{W}^{-1}(\mathbf\Psi,\nu)$, 
if its inverse $\mathbf{X}^{-1}$ has a 
Wishart distribution $\mathcal{W}(\mathbf \Psi^{-1}, \nu)$.