---
title: "My Reproducible Manuscript"
author: "Adam Maghout"
format: pdf: default
bibliography: references.bib
---

ADAM!!! renv has been initialised. MAKE SURE YOU TAKE A SNAPSHOT USING renv::snapshot()

# Purpose of the project

The project presented here is a reproduction of work by @boulesteixIntroductionStatisticalSimulations2020, done for the purpose of an exercise of the Markup Languages and Reproducible Programming in Statistics course offered by Utrecht University. The code used here was provided by the original researchers in pdf form then converted into workable R code by Gerko Vink and Hanne Oberman. Likewise, the data used was also extracted from the website of the [Centers for Disease Control and Prevention (CDC)](https://wwwn.cdc.gov/nchs/nhanes/continuousnhanes/default.aspx?Begi-nYear=2015) by these two researchers. Presenting the findings and formatting this document were done by Adam Maghout.

# Libraries

The required libraries for this project are given below:

```{r libraries, message=F, warning=F}
library(Hmisc)
library(mice)
library(tidyverse)
```

# Data

```{r full code}
# The data can be downloaded in xpt form from https://wwwn.cdc.gov/nchs/nhanes/continuousnhanes/default.aspx?BeginYear=2015
# read data:
d1 <- sasxport.get("../data/DEMO_I.xpt")
d2 <- sasxport.get("../data/BPX_I.xpt")
d3 <- sasxport.get("../data/BMX_I.xpt")
d4 <- sasxport.get("../data/GHB_I.xpt")
d5 <- sasxport.get("../data/TCHOL_I.xpt")
d1.t <- subset(d1,select=c("seqn","riagendr","ridageyr"))
d2.t <- subset(d2,select=c("seqn","bpxsy1"))
d3.t <- subset(d3,select=c("seqn","bmxbmi"))
d4.t <- subset(d4,select=c("seqn","lbxgh"))
d5.t <- subset(d5,select=c("seqn","lbdtcsi"))
d <- merge(d1.t,d2.t)
d <- merge(d,d3.t)
d <- merge(d,d4.t)
d <- merge(d,d5.t)
# ====================================================================
# rename variables:
# RIAGENDR - Gender
# RIDAGEYR - Age in years at screening
# BPXSY1 - Systolic: Blood pres (1st rdg) mm Hg
# BMXBMI - Body Mass Index (kg/m**2)
# LBDTCSI - Total Cholesterol (mmol/L)
# LBXGH - Glycohemoglobin (%)
d$age <- d$ridageyr
d$sex <- d$riagendr
d$bp <- d$bpxsy1
d$bmi <- d$bmxbmi
d$HbA1C <- d$lbxgh
d$chol <- d$lbdtcsi
d$age[d$age<18] <- NA
# ====================================================================
# select complete cases:
dc <- cc(subset(d,select=c("age","sex","bmi","HbA1C","bp")))
# analysis:
summary(lm(bp ~ HbA1C + age + as.factor(sex), data=dc))
confint(lm(bp ~ HbA1C + age + as.factor(sex), data=dc))
summary(lm(bp ~ HbA1C + bmi + age + as.factor(sex), data=dc))
confint(lm(bp ~ HbA1C + bmi + age + as.factor(sex), data=dc))
# ====================================================================
# simulation of measurement error:
ref <- lm(bp ~ HbA1C + bmi + age + as.factor(sex), data=dc)$coef[2]
n.sim <- 1e3
perc.me.exp <- seq(0,.5,.1)
perc.me.conf<- seq(0,.5,.1)
scenarios <- expand.grid(perc.me.exp,perc.me.conf)
var.exp <- var(dc$HbA1C)
var.conf <- var(dc$bmi)
n <- dim(dc)[1]
beta.hat <- matrix(ncol=dim(scenarios)[1], nrow=n.sim)
for (k in 1:n.sim){
  print(k)
  set.seed(k)
  for (i in 1:dim(scenarios)[1]){
    var.me.exp <- var.exp*scenarios[i,1]/(1-scenarios[i,1])
    var.me.conf <- var.conf*scenarios[i,2]/(1-scenarios[i,2])
    dc$HbA1C.me <- dc$HbA1C + rnorm(dim(dc)[1], 0, sqrt(var.me.exp) )
    dc$bmi.me <- dc$bmi + rnorm(dim(dc)[1], 0, sqrt(var.me.conf) )
    beta.hat[k,i] <- lm(bp ~ HbA1C.me + age + bmi.me + as.factor(sex), data=dc)$coef[2]
  }}
# ====================================================================
# create figure:
tot.mat <- cbind(100*scenarios,apply(beta.hat,2,mean))
colnames(tot.mat) <- c("me.exp","me.conf","estimate")
FIGURE <- ggplot(tot.mat, aes(me.exp, me.conf)) +
  geom_tile(color="white",aes(fill = estimate)) +
  geom_text(aes(label = round(estimate, 2))) +
  scale_fill_gradient2(low="#D55E00",mid="white",high = "#56B4E9", midpoint=ref) +
  labs(x=paste("% of total variance of HbA1c due to measurement error"),
       y=paste("% of total variance of BMI due to measurement error")) +
  coord_equal()+
  scale_y_continuous(breaks=unique(tot.mat[,1]))+
  scale_x_continuous(breaks=unique(tot.mat[,1]))+
  theme(panel.background = element_rect(fill='white', colour='grey'),
        plot.title=element_text(hjust=0),
        axis.ticks=element_blank(),
        axis.title=element_text(size=12),
        axis.text=element_text(size=10),
        legend.title=element_text(size=12),
        legend.text=element_text(size=10))
FIGURE
# savePlot("Figure_STRATOS.tif", type="tif")
# ====================================================================
# END OF R CODE
# ===================================================================

```


